---
title: "Structuring the Markov Model: Group Exercise"
format: revealjs
editor: visual
---

## Model Diagram

::: column-page
```{dot}
//| fig-width: 10
strict digraph {
    H [label="Asympomatic (A)"]
    H -> H
    H -> S
    S [label="Progressive Disease (P)"]
    S -> S
    H -> Db
    Db [label="Death from other causes (Db)"]
    S -> Db
    Db -> Db
    S -> Dd
    Dd [label="Death from disease (Dd)"]
    Dd -> Dd
}
```
:::

## Natural History Parameters

-   Progressive disease incidence rate (`r_AP`): 0.4 over 10 years
-   Background mortality rate (`r_Db`): 0.006 per year
-   Hazard rate of death from progressive disease relative to background mortality (`HR_PDd`): 3.0

## Strategy A: Preventive Measure

-   Relative risk of developing progressive disease (`RR_AP`): 0.75

## Strategy B: Disease Treatment

-   Odds-ratio of death from progressive disease (`OR_PDd`): 0.80
-   Baseline annual probability of death in underlying clinical trial (`p0_OR_PDd`): 0.03

## Strategy C: Cure

-   10-year disease remission rate (i.e., return to asymptomatic state; `r_PA_trtC`): 0.50

## Parameter Table

::: {style="font-size: 0.8em"}
| Parameter   | Description                                                                    | Value |
|------------|-------------------------------------------------|------------|
| `r_AP`      | 10-year progressive disease incidence rate                                     | 0.40  |
| `r_Db`      | Background mortality rate                                                      | 0.006 |
| `HR_PDd`    | Hazard rate of death from progressive disease relative to background mortality | 3.0   |
| `RR_AP`     | Relative risk of developing progressive disease (**Strategy A**)               | 0.75  |
| `OR_PDd`    | Odds-ratio of death from progressive disease (**Strategy B**)                  | 0.80  |
| `p0_OR_PDd` | Baseline probability of death in underlying clinical trial (**Strategy B**)    | 0.03  |
| `r_PA_trtC` | 10-year disease remission rate (**Strategy C**)                                | 0.50  |
:::

```{r}
library(tidyverse)
library(Matrix)
r_AP <- 0.40
r_Db <- 0.006
HR_PDd <- 3.0
RR_AP <- 0.75
OR_PDd <- 0.80
p0_OR_PDd <- 0.03
r_PA_trtC <- 0.50

m_R0 <- m_RA <- m_RB <- m_RC <-   
  matrix(0, nrow=4, ncol = 4, dimnames=list(c("Asymptomatic","Progressive Disease","Disease Mortality", "Background Mortality"),
                                               c("Asymptomatic","Progressive Disease","Disease Mortality", "Background Mortality")))

# Step 1: Natural History  (i.e., m_R0 and m_P0)

m_R0["Asymptomatic","Asymptomatic"] <- -(r_AP/10+r_Db)
m_R0["Asymptomatic","Progressive Disease"] <- r_AP/10
m_R0["Asymptomatic","Background Mortality"] <- r_Db

m_R0["Progressive Disease","Progressive Disease"] <- -(HR_PDd * r_Db + r_Db)
m_R0["Progressive Disease","Disease Mortality"] <- HR_PDd * r_Db
m_R0["Progressive Disease","Background Mortality"] <- r_Db

m_P0 <- expm(m_R0)
m_P0[is.na(m_P0)] <- 0

# Strategy A

m_PA <- m_P0 
m_PA["Asymptomatic","Progressive Disease"] <- RR_AP * m_P0["Asymptomatic","Progressive Disease"] 
m_PA["Asymptomatic","Asymptomatic"] <- 1 - m_PA["Asymptomatic","Progressive Disease"]  - m_PA["Asymptomatic","Disease Mortality"] - m_PA["Asymptomatic","Background Mortality"]

# Strategy B
m_PB <- m_P0 
m_PB["Progressive Disease","Disease Mortality"] <- (OR_PDd / (1 - p0_OR_PDd + (p0_OR_PDd *OR_PDd  ))) * m_PB["Progressive Disease","Disease Mortality"]
m_PB["Progressive Disease","Progressive Disease"] <- 1 - m_PB["Progressive Disease","Asymptomatic"] - m_PB["Progressive Disease","Disease Mortality"] - m_PB["Progressive Disease","Background Mortality"]

# Strategy C

m_RC <- m_R0 
m_RC["Progressive Disease","Asymptomatic"]  <- (r_PA_trtC/10)
m_RC["Progressive Disease","Progressive Disease"] <- -(m_RC["Progressive Disease","Asymptomatic"] + m_RC["Progressive Disease","Disease Mortality"] + m_RC["Progressive Disease","Background Mortality"])
m_PC <- expm(m_RC)

```

## Key

|                                                                   |
|-------------------------------------------------------------------|
| </span>[Rate adjusted for timestep]{style="background:#ff498c"}   |
| </span>[Hazard Ratio]{style="background:#fcb08a"}                 |
| </span>[Probability]{style="background:#ffc922"}                  |
| </span>[Relative Risk]{style="background:#cdea24"}                |
| </span>[Odds Ratio]{style="background:#9dd8da"}                   |
| </span>[Constructed from other cells]{style="background:#ab59cf"} |

```{r}

color_lut <- 
  c("rate" = "#ff498c",
    "hazard_ratio" = "#fcb08a",
    "probability" = "#ffc922",
    "relative_risk" = "#cdea24", 
    "odds_ratio" = "#9dd8da", 
    "constructed" = "#ab59cf")

```

## Goal: Natural History Transition Probability Matrix

**m_P0** =

::: {style="font-size: 0.5em"}
|                      | Asymptomatic                                                         | Progressive Disease                                                  | Disease Mortality                                                    | Background Mortality                                                 |
|---------------|---------------|---------------|---------------|---------------|
| Asymptomatic         | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ab59cf"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} |
| Progressive Disease  | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ab59cf"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} |
| Disease Mortality    | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ab59cf"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} |
| Background Mortality | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ffc922"} | </span>[$\quad \quad \quad \quad \quad$]{style="background:#ab59cf"} |
:::

::: {style="font-size: 0.6em"}
|                                                                   |
|-------------------------------------------------------------------|
| </span>[Rate adjusted for timestep]{style="background:#ff498c"}   |
| </span>[Hazard Ratio]{style="background:#fcb08a"}                 |
| </span>[Probability]{style="background:#ffc922"}                  |
| </span>[Relative Risk]{style="background:#cdea24"}                |
| </span>[Odds Ratio]{style="background:#9dd8da"}                   |
| </span>[Constructed from other cells]{style="background:#ab59cf"} |
:::
