---
title: "What They Didn't Teach You About Decision Modeling: CVD Model"
output:
  learnr::tutorial:
    progressive: true
    ace_theme: github
    theme: united
runtime: shiny_prerendered
description: >
  This tutorial will cover background mortality modeling and model construction for a simple cardiovascular disease natural history model".
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(demography)
library(MortalityLaws)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_chunk$set(echo = FALSE)
#knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
#set_here(here::here("_learnr/smdm-europe-2023-cvd-model/"))
```

## Part 1: Mortality Modeling

In this section we will explore mortality modeling using the `demography` and `MortalityLaws` packages, and using data from the Human Mortality Database

```{r, eval = FALSE}
hmd.usa <- demography::hmd.mx("USA",username = "", password = "", "USA")
write_rds(hmd.usa,file=here("_learnr/smdm-europe-2023-cvd-model/www/usa-life-table.rds"))
```

Our first objective is to download and load the Human Mortality Database life table data.

```{r}
#| eval: false
#| echo: true
hmd_usa <- demography::hmd.mx("USA",username = "<<your user name>>", password = "<<your password>>", "USA")
```

We next use the `demography::lifetable()` to extract the life table data for a single year (2019) and for a population radix (i.e., starting population) of 100,000 indiviudals. 

```{r}
#| eval: false
#| echo: true

mortality_year = 2019
radix = 100000

lt = 
  hmd_usa %>% 
  demography::lifetable(.,series = "total", years = mortality_year) %>% 
  as_tibble() %>% 
  mutate_at(vars(lx,dx), function(x) x * radix) %>% 
  mutate(country = "USA") %>% 
  mutate(age = x)

```

Let's take a look at our life table data:

```{r loadlt}

mortality_year = 2019
radix = 100000

lt <- 
  list.files(recursive=TRUE,full.names = TRUE) %>% 
  grep("www/usa-life-table.rds",.,value=TRUE) %>% 
  read_rds(.) %>% 
  demography::lifetable(.,series = "total", years = mortality_year) %>% 
  as_tibble() %>% 
  mutate_at(vars(lx,dx), function(x) x * radix) %>% 
  mutate(country = "USA") %>% 
  mutate(age = x)

lt %>% 
  ungroup() %>% 
  select(-x) %>% 
  select(country,age,everything()) %>% 
  head() %>% 
  kable() %>% 
  kable_styling()

```

The columns here are 

- `age`:	Ages for lifetable
- `year`: Period years or cohort years
- `mx`: 	Death rate at age x.
- `qx`: The probability that an individual of exact age x will die before exact age x+1.
- `lx`: Number of survivors to exact age x. This is defined relative to a **radix**, or the size of a cohort from which the life table is derived).
- `dx`: The number of deaths between exact ages x and x+1.
- `Lx`: Number of years lived between exact age x and exact age x+1.
- `Tx`: Number of years lived after exact age x.
- `ex`: Remaining life expectancy at exact age x.

## Parametric Mortality Models

Our next step is to fit a mortality model to these data. Generally speaking, we need three inputs:

- `age`:	Ages for lifetable
- `dx`: The number of deaths between exact ages x and x+1.
- `lx`: Number of survivors to exact age x. This is defined relative to a **radix**, or  the size of a cohort from which the life table is derived).

The `MortalityLaws` package has a *number* of mortality models we can draw from: 

```{r}
#| tbl-cap: Mortality Models in MortalityLaws package

type_lut <- c(
"1"  =  "Infant mortality" ,             
"2"  =  "Accident hump"      ,           
"3"  =  "Adult mortality"   ,           
"4"  =  "Adult and/or old-age mortality",
"5"  =  "Old-age mortality"       ,      
"6"  =  "Full age range" 
)
 
laws <- 
  availableLaws()$table %>% 
  as_tibble() %>% 
  arrange(TYPE) %>% 
  mutate(TYPE = type_lut[paste0(TYPE)]) %>% 
  #filter(CODE=="HP2") %>% 
  #mutate(MODEL = gsub("\\^","||",MODEL))
  mutate(MODEL = paste0("$",MODEL,"$")) %>% 
  mutate(MODEL = ifelse(CODE == "weibull","$mu[x] = 1/sigma * (x/M)^{(M/sigma - 1)}$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "invweibull","$mu[x] = 1/sigma * (x/M)^{[-M/sigma - 1]} / [exp((x/M)^{(-M/sigma)}) - 1]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "perks","$mu[x] = [A + BC^x] / [BC^{-x} + 1 + DC^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "wittstein","$q[x] = (1/B) A^{-[(Bx)^N]} + A^{-[(M-x)^N]}$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP","$q[x]/p[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + G H^x$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP2","$q[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + GH^x / [1 + GH^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP3","$q[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + GH^x / [1 + KGH^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP3","$q[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + GH^x / [1 + KGH^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "kostaki","$q[x]/p[x] = A^{[(x+B)^C]} + D exp[-(E_i log(x/F_))^2] + G H^x$",MODEL))
  

laws %>% 
  select(-YEAR,-TYPE,-FIT,-SCALE_X) %>% 
  kable(col.names = c("Name","Model","Code"))# %>% 
  #pack_rows(index = table(fct_inorder(laws$TYPE)))
```

## Fitting a Parametric Mortality Model

```{r define, exercise = TRUE, exercise.setup = "loadlt"}


lt

```


## Part 2: Alive-Dead Model

## Part 3: Healthy-CVD-Dead Model

## Part 4: Advanced Bookkeeping 



