---
title: "1. Mortality Modeling"
output:
  learnr::tutorial:
    progressive: true
    theme: united
    allow_skip: true
runtime: shiny_prerendered
description: >
  This tutorial will cover background mortality modeling".
---

```{r setup, include=FALSE}
# ("rstudio", default", "cerulean", "journal", "flatly", "readable",
# "spacelab", "united", "cosmo", "lumen", "paper", "sandstone", "simplex", or
# "yeti").

    # ace_theme: github
    # theme: united
library(learnr)
library(tidyverse)
library(demography)
library(MortalityLaws)
library(directlabels)
library(ggsci)
library(hrbrthemes)
library(MASS)
library(mgcv)
library(patchwork)
library(knitr)
library(kableExtra)
library(here)
library(ggsci)
library(expm)
select <- dplyr::select
options("scipen" = 100, "digits" = 5)

knitr::opts_chunk$set(echo = FALSE)
options("scipen" = 100, "digits" = 5)
#knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
#set_here(here::here("_learnr/smdm-europe-2023-cvd-model/"))
```

In this section we will explore mortality modeling using the `demography` and `MortalityLaws` packages, and using data from the Human Mortality Database

```{r, eval = FALSE}
hmd.usa <- demography::hmd.mx("USA",username = "", password = "", "USA")
write_rds(hmd.usa,file=here("_learnr/smdm-europe-2023-cvd-model/www/usa-life-table.rds"))
```

Our first objective is to download and load the Human Mortality Database life table data.

```{r}
#| eval: false
#| echo: true
hmd_usa <- demography::hmd.mx("USA",username = "<<your user name>>", password = "<<your password>>", "USA")
```

We next use the `demography::lifetable()` to extract the life table data for a single year (2019) and for a population radix (i.e., starting population) of 100,000 indiviudals.

```{r}
#| eval: false
#| echo: true

mortality_year = 2019
radix = 100000

lt = 
  hmd_usa %>% 
  demography::lifetable(.,series = "total", years = mortality_year) %>% 
  as_tibble() %>% 
  mutate_at(vars(lx,dx), function(x) x * radix) %>% 
  mutate(country = "USA") %>% 
  mutate(age = x)

```

Let's take a look at our life table data:

```{r}
mortality_year = 2019
radix = 100000

lt_usa_file <- "https://github.com/graveja0/SMDM-Europe-2023/raw/main/_learnr/smdm-europe-2023-cvd-model/www/usa-life-table.rds"

lt <- 
  readRDS(url(lt_usa_file)) %>% 
  demography::lifetable(.,series = "total", years = mortality_year) %>% 
  as_tibble() %>% 
  mutate_at(vars(lx,dx), function(x) x * radix) %>% 
  mutate(country = "USA") %>% 
  mutate(age = x)
```

```{r}
lt %>%
  ungroup() %>%
  select(-x) %>%
  select(country,age,everything()) %>%
  head() %>%
  kable() %>%
  kable_styling()
```

The columns here are

-   `age`: Ages for lifetable
-   `year`: Period years or cohort years
-   `mx`: Death rate at age x.
-   `qx`: The probability that an individual of exact age x will die before exact age x+1.
-   `lx`: Number of survivors to exact age x. This is defined relative to a **radix**, or the size of a cohort from which the life table is derived).
-   `dx`: The number of deaths between exact ages x and x+1.
-   `Lx`: Number of years lived between exact age x and exact age x+1.
-   `Tx`: Number of years lived after exact age x.
-   `ex`: Remaining life expectancy at exact age x.

### Parametric Mortality Models

Our next step is to fit a mortality model to these data. Generally speaking, we need three inputs:

-   `age`: Ages for lifetable
-   `dx`: The number of deaths between exact ages x and x+1.
-   `lx`: Number of survivors to exact age x. This is defined relative to a **radix**, or the size of a cohort from which the life table is derived).

The `MortalityLaws` package has a *number* of mortality models we can draw from:

```{r}
#| tbl-cap: Mortality Models in MortalityLaws package

type_lut <- c(
"1"  =  "Infant mortality" ,             
"2"  =  "Accident hump"      ,           
"3"  =  "Adult mortality"   ,           
"4"  =  "Adult and/or old-age mortality",
"5"  =  "Old-age mortality"       ,      
"6"  =  "Full age range" 
)
 
laws <- 
  availableLaws()$table %>% 
  as_tibble() %>% 
  arrange(TYPE) %>% 
  mutate(TYPE = type_lut[paste0(TYPE)]) %>% 
  #filter(CODE=="HP2") %>% 
  #mutate(MODEL = gsub("\\^","||",MODEL))
  mutate(MODEL = paste0("$",MODEL,"$")) %>% 
  mutate(MODEL = ifelse(CODE == "weibull","$mu[x] = 1/sigma * (x/M)^{(M/sigma - 1)}$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "invweibull","$mu[x] = 1/sigma * (x/M)^{[-M/sigma - 1]} / [exp((x/M)^{(-M/sigma)}) - 1]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "perks","$mu[x] = [A + BC^x] / [BC^{-x} + 1 + DC^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "wittstein","$q[x] = (1/B) A^{-[(Bx)^N]} + A^{-[(M-x)^N]}$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP","$q[x]/p[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + G H^x$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP2","$q[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + GH^x / [1 + GH^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP3","$q[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + GH^x / [1 + KGH^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "HP3","$q[x] = A^{[(x + B)^C]} + D exp[-E log(x/F)^2] + GH^x / [1 + KGH^x]$",MODEL)) %>% 
  mutate(MODEL = ifelse(CODE == "kostaki","$q[x]/p[x] = A^{[(x+B)^C]} + D exp[-(E_i log(x/F_))^2] + G H^x$",MODEL))
  

laws %>% 
  select(-YEAR,-TYPE,-FIT,-SCALE_X) %>% 
  kable(col.names = c("Name","Model","Code")) %>% 
  pack_rows(index = table(fct_inorder(laws$TYPE)))
```

### Fitting a Parametric Mortality Model, Part 1

```{r loadlt}
mortality_year = 2019
radix = 100000

lt_usa_file <- "https://github.com/graveja0/SMDM-Europe-2023/raw/main/_learnr/smdm-europe-2023-cvd-model/www/usa-life-table.rds"

lt <- 
  readRDS(url(lt_usa_file)) %>% 
  demography::lifetable(.,series = "total", years = mortality_year) %>% 
  as_tibble() %>% 
  mutate_at(vars(lx,dx), function(x) x * radix) %>% 
  mutate(country = "USA") %>% 
  mutate(age = x)
```

Use one of the mortality laws to fit a parametric model to the life table data. You can fit this model over any age range (e.g., 0-99, 40-99, etc.).

```{r parmort, exercise = TRUE, exercise.eval = FALSE, exercise.setup="loadlt"}

min_age = 0
max_age = 99

ages     <- lt$x[lt$x<=max_age & lt$x>=min_age]
deaths   <- lt$dx[lt$x<=max_age & lt$x>=min_age]
exposure <- lt$lx[lt$x<=max_age & lt$x>=min_age]

mort_fit <- MortalityLaw(
                x  = ages,
                Dx  = deaths,   # vector with death counts
                Ex  = exposure, # vector containing exposures
                law = "",
                opt.method = "LF2")
plot(mort_fit)
```

### Fitting a Parametric Mortality Model, Part 2

Each mortality law has its own function to predict the mortality rate at a given time. In the block below, fit a model for a starting cohort of 40 year olds using the `HP2` mortality law, and predict the mortality rate for a 75 year old.

The basic input into the mortality law function is `x`, or the time period, and the coefficients from the fitted model object.

```{r cohort40, exercise = TRUE, exercise.eval = FALSE, exercise.setup="loadlt"}
#| exercise.lines: 20
#| 
min_age = 
max_age = 99

ages     <- lt$x[lt$x<=max_age & lt$x>=min_age]
deaths   <- lt$dx[lt$x<=max_age & lt$x>=min_age]
exposure <- lt$lx[lt$x<=max_age & lt$x>=min_age]

mort_fit <- MortalityLaw(
                x  = ages,
                Dx  = deaths,   # vector with death counts
                Ex  = exposure, # vector containing exposures
                law = "",
                opt.method = "LF2")

r_mort75 <- 
```

```{r cohort40-solution}
min_age = 40
max_age = 99

ages     <- lt$x[lt$x<=max_age & lt$x>=min_age]
deaths   <- lt$dx[lt$x<=max_age & lt$x>=min_age]
exposure <- lt$lx[lt$x<=max_age & lt$x>=min_age]

mort_fit <- MortalityLaw(
                x  = ages,
                Dx  = deaths,   # vector with death counts
                Ex  = exposure, # vector containing exposures
                law = "HP2",
                opt.method = "LF2")


r_mort75 <-  HP2(x = 75 - min_age, coef(mort_fit))
r_mort75
```
